---
title: "Regression models"
output: html_document
date: "`r Sys.Date()`"
---

# Data inputting and libraries

```{r}
rm(list = ls())
getwd()
setwd("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project")
df_long <- read.csv("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project/Data/all_data_long.csv")
df_wide <- read.csv("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project/Data/all_data_wide.csv")

df_long <- as.data.frame(df_long)
df_wide <- as.data.frame(df_wide)

library(dplyr)
library(plm)
library(foreign)
library(psych)
library(lmtest)
library(tseries)
library(stargazer)
library(car)
library(sandwich)
```

# Shapiro-Wilk test for normality of distribution

### As expected, distributions are not normal

```{r}
shapiro_by_country <- function(df) {
  
  result_list <- list()
  variables <- setdiff(names(df), c("Date", "Country"))
  
  # Loop through each country
  for (country in unique(df$Country)) {
    country_data <- df %>% filter(Country == country)
    
    # Loop through each variable
    for (variable in variables) {
      valid_data <- country_data[[variable]]  
      valid_data <- valid_data[!is.na(valid_data)]  
      
      if(length(valid_data) > 3) {  
        test_result <- shapiro.test(valid_data)
        result_list[[paste(country, variable, sep = "_")]] <- test_result$p.value
      } else {
        result_list[[paste(country, variable, sep = "_")]] <- NA
      }
    }
  }
  
  result_df <- as.data.frame(result_list)
  result_df <- t(result_df) 
  colnames(result_df) <- "p_value"
  
  return(result_df)
}

shapiro_results <- shapiro_by_country(df_wide)
print(shapiro_results)
```

# Regressions

## Pooled OLS

```{r}
pooled <- plm(crime_rate ~ exp_years_of_school  + unemployment_ + gini_coef + GDP, 
              data = df_wide,
              model = "pooling")
summary(pooled)
```

## Fixed Effects model

```{r}
fe1 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school + unemployment_ + gini_coef  + GDP, 
           data = df_wide, 
           index = c("Country", "Date"), 
           model = "within")
summary(fe1)

fe2 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school  + gini_coef + GDP, 
          data = df_wide, 
          index = c("Country", "Date"), 
          model = "within")
summary(fe2)

fe3 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school  + GDP, 
           data = df_wide, 
           index = c("Country", "Date"), 
           model = "within")
summary(fe3)
```

### Fe2 is the best model

# Handling autocorrelation - two methods: dummy Year and lags

## factor(Date)

```{r}
fey1 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school + unemployment_ + gini_coef + GDP + factor(Date), 
            data = df_wide, 
            index = c("Country", "Date"), 
            model = "within")
summary(fey1)

fey2 <- plm(crime_rate ~ exp_years_of_school + gini_coef + GDP + factor(Date), 
            data = df_wide, 
            index = c("Country", "Date"), 
            model = "within")
summary(fey2)

fey3 <- plm(crime_rate ~ exp_years_of_school + GDP + factor(Date), 
            data = df_wide, 
            index = c("Country", "Date"), 
            model = "within")
summary(fey3)
```

## Introducing Lags

```{r}
fet1 <- plm(crime_rate ~ lag(crime_rate, 1) + exp_years_of_school   + gini_coef + GDP, 
           data = df_wide, 
           index = c("Country", "Date"), 
           model = "within")
summary(fet1)

fet2 <- plm(crime_rate ~ lag(crime_rate, 1) + exp_years_of_school + gini_coef  + GDP + lag(GDP, 1), 
            data = df_wide, 
            index = c("Country", "Date"), 
            model = "within")
summary(fet2)

fet3 <- plm(crime_rate ~ lag(crime_rate, 1) + exp_years_of_school + gini_coef + lag(gini_coef, 1) + GDP + lag(GDP, 1), 
            data = df_wide, 
            index = c("Country", "Date"), 
            model = "within")
summary(fet3)
```

## FD model - just for checking other possibilities

```{r}
fd1 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school + unemployment_ + gini_coef + GDP, 
           data = df_wide, 
           index = c("Country", "Date"), 
             model = "fd")
summary(fd1)
fd2 <- plm(crime_rate ~ exp_years_of_school + mean_years_of_school + gini_coef + GDP , 
           data = df_wide, 
           index = c("Country", "Date"), 
           model = "fd")
summary(fd2)
```

# Generating residuals vs fitted values plots

### Changing fe1 to other models will provide plots for each model

```{r}
fitted_fe1 <- as.numeric(fitted(fe1))
residuals_fe1 <- as.numeric(residuals(fe1))

png("residual_plot_fe1.png", width = 700, height = 300)
plot(fitted_fe1, residuals_fe1, 
     main = "Residuals vs Fitted Values (FE with dummy Year)", 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     pch = 16)  
abline(h = 0, col = "red", lwd = 2)  
dev.off()
```

# TESTS

## Poolability test

```{r}
z = pvcm(crime_rate ~ exp_years_of_school + mean_years_of_school + unemployment_ + gini_coef + GDP, 
         data = df_wide, 
         index = c("Country", "Date"), 
         model = "within")

pooltest(fe1, z)
```

## Collinearity

```{r}
X <- model.matrix(fet2)
eigenvals <- eigen(cor(X))$values
condition_index <- sqrt(max(eigenvals)/eigenvals)
condition_index
```

## Autocorrelation - Breusch - Godfrey test

```{r}
pbgtest(fet2)
```

## Heteroscedasticity - Breusch - Pagan test

```{r}
bptest(crime_rate ~ lag(crime_rate, 1) + exp_years_of_school  + gini_coef + GDP + lag(GDP, 1), 
       data = df_wide)
```

## Introducing Robust standard errors -\> (HAC) covariance matrix

```{r}
coeftest(fet2, vcov = vcovHC(fet2, type = "HC1"))

fet2$vcov <- vcovHC(fet2, type = "HC1")
```

## Cross-sectional Dependence - Pesaran's Test

```{r}
pcdtest(fet2)
```

# Stargazer for table of models

```{r}
stargazer(fe1, fe2, fet1, fet2)
```
