---
title: "Data Exploration"
output: html_document
date: "`r Sys.Date()`"
---

# Loading libraries and inputting data

```{r setup, include=FALSE}



rm(list = ls())
library(plm)
library(foreign)
library(psych)
library(ggplot2)
library(data.table)
library(lmtest)
library(corrplot)
library(knitr)
library(kableExtra)
library(dplyr)
library(gridExtra)

getwd()
setwd("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project")
df_long <- read.csv("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project/Data/all_data_long.csv")
df_wide <- read.csv("C:/Users/Stanislav/OneDrive/Pulpit/Szwajcaria/Studies/Empirical Methods in Economics/Project/Data/all_data_wide.csv")

df_long <- as.data.frame(df_long)
df_wide <- as.data.frame(df_wide)
```

# Data Exploration

## Summary statistics

```{r}
# Custom summary function that includes standard deviation for data frames
summary_with_sd_df <- function(df) {
  df_numeric <- df[sapply(df, is.numeric)]
  
  summary_stats <- lapply(df_numeric, function(x) {
    min_val <- min(x, na.rm = TRUE)
    q1 <- quantile(x, 0.25, na.rm = TRUE)
    median_val <- median(x, na.rm = TRUE)
    mean_val <- mean(x, na.rm = TRUE)
    q3 <- quantile(x, 0.75, na.rm = TRUE)
    max_val <- max(x, na.rm = TRUE)
    sd_val <- sd(x, na.rm = TRUE)
    n_missing <- sum(is.na(x))
    
    c(
      "Min." = min_val,
      "1st Qu." = q1,
      "Median" = median_val,
      "Mean" = mean_val,
      "3rd Qu." = q3,
      "Max." = max_val,
      "Std Dev" = sd_val,
      "NA's" = n_missing
    )
  })
  
  summary_df <- do.call(cbind, summary_stats)
  colnames(summary_df) <- colnames(df_numeric)
  return(summary_df)
}

df_val <- df_wide[3:8]

summary_stats <- summary_with_sd_df(df_val)
kable(summary_stats, digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Summary Statistics for each variable for each country

### Subsetting data for each country

```{r}
df_germany <- df_wide %>%
  filter(Country == 'germany')
df_us <- df_wide %>%
  filter(Country == 'us')
df_uk <- df_wide %>%
  filter(Country == 'uk')
df_canada <- df_wide %>%
  filter(Country == 'canada')
df_costa_rica <- df_wide %>%
  filter(Country == 'costa_rica')
df_brazil <- df_wide %>%
  filter(Country == 'brazil')
```

### Removing Country columns

```{r}
df_germany$Country <- NULL
df_us$Country <- NULL
df_uk$Country <- NULL
df_canada$Country <- NULL
df_costa_rica$Country <- NULL
df_brazil$Country <- NULL
```

### Individual summary statistics

```{r}
summary_stats_germany <- summary_with_sd_df(df_germany %>% select(-Date))
write.csv(summary_stats_germany, "summary_stats_germany.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for Germany") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

summary_stats_us <- summary_with_sd_df(df_us %>% select(-Date))
write.csv(summary_stats_us, "summary_stats_us.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for United States") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

summary_stats_uk <- summary_with_sd_df(df_uk %>% select(-Date))
write.csv(summary_stats_uk, "summary_stats_uk.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for United Kingdom") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

summary_stats_canada <- summary_with_sd_df(df_canada %>% select(-Date))
write.csv(summary_stats_canada, "summary_stats_canada.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for Canada") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

summary_stats_costa_rica <- summary_with_sd_df(df_costa_rica %>% select(-Date))
write.csv(summary_stats_costa_rica, "summary_stats_costa_rica.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for Costa Rica") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

summary_stats_brazil <- summary_with_sd_df(df_brazil %>% select(-Date))
write.csv(summary_stats_brazil, "summary_stats_brazil.csv", row.names = TRUE)
kable(summary_stats, digits = 3, caption = "Descriptive statistics for Brazil") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Plots

## Filtering variables

```{r}
df_crime_rate <- df_long %>%
  filter(Indicator == "crime_rate")
df_exp_school <- df_long %>%
  filter(Indicator == "exp_years_of_school")
df_mean_school <- df_long %>%
  filter(Indicator == "mean_years_of_school")
df_gini <- df_long %>%
  filter(Indicator == "gini_coef")
df_unemployment <- df_long %>%
  filter(Indicator == "unemployment_")
df_gdp <- df_long %>%
  filter(Indicator == 'GDP')
```

## Time plots of variables

```{r}
ggplot(data = df_crime_rate, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Crime Rate Trends by Country",
    x = "Year",
    y = "Intentional homicides per 100k population"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  ) 
ggsave("crime_rate_plot.png", width = 8, height = 6, dpi = 300)  


ggplot(data = df_exp_school, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Expected number of years of education by Country",
    x = "Year",
    y = "Expected years of school education"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  ) 
ggsave("exp_years_of_school_plot.png", width = 8, height = 6, dpi = 300) 


ggplot(data = df_mean_school, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Mean numbers of years of education by Country",
    x = "Year",
    y = "Mean number of years of education"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  ) 
ggsave("mean_years_of_school_plot.png", width = 8, height = 6, dpi = 300)


ggplot(data = df_unemployment, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Unemployment by Country",
    x = "Year",
    y = "Unemployment in %"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  ) 
ggsave("unemployment_plot.png", width = 8, height = 6, dpi = 300)


ggplot(data = df_gini, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "Gini Coefficient by Country",
    x = "Year",
    y = "Gini Coefficient"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  )
ggsave("gini_plot.png", width = 8, height = 6, dpi = 300)

ggplot(data = df_gdp, aes(x = Date, y = Value, color = Country, group = Country)) +
  geom_line(size = 1) +
  labs(
    title = "GDP per capita by Country",
    x = "Year",
    y = "GDP per capita (USD)"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 20, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13)
  ) 
ggsave("gdp_per_cap_plot.png", width = 8, height = 6, dpi = 300)
```

## Scatterplots

```{r}
# crime and exp
s1 <- ggplot(df_wide, aes(x = exp_years_of_school, y = crime_rate, color = Country)) +
  geom_point() +
  labs(title = "Scatter Plot of Expected Education vs Crime Rate",
       x = "Years of Expected Education",
       y = "Crime Rate",
       color = "Country")  +
  theme_minimal()+
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13),
    legend.position = "None"
  )#+
 #geom_smooth(method = "lm", aes(group = Country), se = FALSE)
#ggsave("scatterplot_crime_exp.png", width = 8, height = 6, dpi = 300)

# crime and mean school
s2 <- ggplot(df_wide, aes(x = mean_years_of_school, y = crime_rate, color = Country)) +
  geom_point() +
  labs(title = "Scatter Plot of Received Education vs Crime Rate",
       x = "Years of Received Education",
       y = "Crime Rate",
       color = "Country")+ 
  theme_minimal()+
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13),
    legend.position = "None"
  )#+
  #geom_smooth(method = "lm", aes(group = Country), se = FALSE)
ggsave("scatterplot_crime_mean.png", width = 8, height = 6, dpi = 300)

# crime and unemployment
s3 <- ggplot(df_wide, aes(x = unemployment_, y = crime_rate, color = Country)) +
  geom_point() +
  labs(title = "Scatter Plot of Unemployment vs Crime Rate",
       x = "Unemployment",
       y = "Crime Rate",
       color = "Country")+ 
  theme_minimal()+
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13),
    legend.position = "None"
  )#+
  #geom_smooth(method = "lm", aes(group = Country), se = FALSE)
#ggsave("scatterplot_crime_unemployment.png", width = 8, height = 6, dpi = 300)

# crime and gini
s4 <- ggplot(df_wide, aes(x = gini_coef, y = crime_rate, color = Country)) +
  geom_point() +
  labs(title = "Scatter Plot of Gini Coefficient vs Crime Rate",
       x = "Gini Coefficient",
       y = "Crime Rate",
       color = "Country")+ 
  theme_minimal()+
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 17),            
    legend.text = element_text(size = 13),
    legend.position = "None"
  )#+
  #geom_smooth(method = "lm", aes(group = Country), se = FALSE)
#ggsave("scatterplot_crime_gini.png", width = 8, height = 6, dpi = 300)

s5 <- ggplot(df_wide, aes(x = GDP, y = crime_rate, color = Country)) +
  geom_point() +
  labs(title = "Scatter Plot of GDP per capita vs Crime Rate",
       x = "GDP per capita",
       y = "Crime Rate",
       color = "Country")+ 
  theme_minimal()+
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),   
    axis.title = element_text(size = 17),               
    axis.text = element_text(size = 12),                
    legend.title = element_text(size = 20),            
    legend.text = element_text(size = 15),
    legend.position = c(1.7, 0.5)
  )

# Scatterplots together
png("scatterplots.png", width = 1000, height = 500)
grid.arrange(s1, s2, s3, s4, s5, ncol = 3)
dev.off()
```

## Histograms

```{r}
p1 <- ggplot(data = df_wide, aes(x = crime_rate, fill = Country)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Crime Rate by Country", x = "Crime Rate", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
p2 <- ggplot(data = df_wide, aes(x = exp_years_of_school, fill = Country)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Expected Years of School by Country", x = "Expected Years of School", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
p3 <- ggplot(data = df_wide, aes(x = mean_years_of_school, fill = Country)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Crime Rate by Country", x = "Mean Years of School", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
p4 <- ggplot(data = df_wide, aes(x = unemployment_, fill = Country)) +
  geom_histogram(binwidth = 0.3, color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Unemployment by Country", x = "Unemployment (%)", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
p5 <- ggplot(data = df_wide, aes(x = GDP, fill = Country)) +
  geom_histogram( color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "GDP per capita by Country", x = "GDP per capita", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
p6 <- ggplot(data = df_wide, aes(x = gini_coef, fill = Country)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Gini Coefficient by Country", x = "Gini Coefficient", y = "Frequency")+
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right",  
        legend.text = element_text(size = 13), 
        legend.title = element_text(size = 15),  
        legend.key.size = unit(1.1, "cm"))


png("histograms.png", width = 1000, height = 800)
grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)
dev.off()
```

## Correlation Matrix Overall

```{r}
panel_df <- pdata.frame(df_wide, index = c("Date", "Country"))
colnames(panel_df) <- c('Date', 'Country', 'Crime Rate', 'Expected Education', 'Mean Education', 'Unemployment', 'Gini', 'GDP per capita')
pdf_no_indx <- panel_df[, -c(1, 2)]
cor_matrix <- cor(na.omit(pdf_no_indx))

print(corr.test(pdf_no_indx))

kable(cor_matrix, caption = "Correlation Matrix of mtcars Dataset")

png("correlation_matrix.png", width = 1000, height = 750)
corrplot(cor_matrix, method = "number", #square, number, circle, shade, pie, color, ellipse
         number.cex = 3, 
         cex.axis = 1.3,
         type = "upper", order = "original",  
         tl.col = "black", 
         tl.cex = 1.2,
         tl.srt = 0, 
         cl.cex = 1,
         diag = F)
title(main = "Overall Correlation Matrix", 
      col.main = "black",     
      cex.main = 3,             
      font.main = 2,            
      line = 2) 
dev.off()
```

## Individual correlation matrices

```{r}

colnames(df_wide) <- c('Date', 'Country', 'Crime Rate', 'Expected Education', 'Mean Education', 'Unemployment', 'Gini', 'GDP per capita')
generate_correlation_plots <- function(df, countries_list, output_dir = "correlation_plots") {
  
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }
  
  for (country in countries_list) {
    
    # Subset the data for the current country
    df_country <- df_wide[df_wide$Country == country, ]
    
    # Convert to pdata.frame for panel data structure
    panel_df_country <- pdata.frame(df_country, index = c("Date", "Country"))
    
    # Drop the index columns for correlation calculation
    pdf_no_indx_country <- panel_df_country[, -c(1, 2)]
    
    # Calculate the correlation matrix
    cor_matrix_country <- cor(na.omit(pdf_no_indx_country))
    
    # Set file name for saving the plot
    file_name <- paste0(output_dir, "/correlation_matrix_", gsub(" ", "_", country), ".png")
    
    # Generate and save the correlation plot for the current country
    png(file_name, width = 1000, height = 750)
    corrplot(cor_matrix_country, method = "number", # Options: square, number, circle, shade, pie, color, ellipse
             number.cex = 3, 
             cex.axis = 1.3,
             type = "upper", order = "original",  
             tl.col = "black", 
             tl.cex = 1.2,
             tl.srt = 0, 
             cl.cex = 1,
             diag = F)
    title(main = paste("Correlation Matrix for", country), 
          col.main = "black",     
          cex.main = 3,             
          font.main = 2,            
          line = 2) 
    dev.off()  # Close the file output
  }
}

countries_list <- unique(df_wide$Country)
generate_correlation_plots(df_wide, countries_list, output_dir = "correlation_plots")

```
